Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2025-01-27 analysis/paper/paper.Rmd
MS revisions in progress

resultScenario = "S10"
resultSet ="cmult6ay1aSf1"
figSet = c("diffslow","distScns","EVSI","power","EVPIall")

@@

To investigate the effectiveness of various monitoring strategies, and to show how this
varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories. To simulated example trajectories we use a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al. [-@dyson_existing_2022]. For brevity, we refer to this as the "national" model. For clarity, we include a full description of the model in the following paragraphs. Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population. To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using a Bayesian model (section 2.4). We also use the sets of simulated trajectories to confirm that prior predictions from our Bayesian model are consistent with knowledge of the observed distribution of outcomes for caribou across the country [@johnson_science_2020].

@@

In our simulated example populations, mortality risk does not vary over
time or among individuals, so the observed state (dead or alive) of a
collared animal $i$ in month $m$ and year $t$ depends on the true
survival rate $\dot{S}_t$ as follows: 
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(\dot{S}^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatI)
\end{equation}

@@

Local demographic and covariate data provided as inputs to the Bayesian
population model include the observed state (dead or alive) of each
collared animal in each month ($\hat{I}_{i,t,m}$), the observed number
of adult females and calves in the recruitment surveys
($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per
collared female in the composition survey $w$, and the measures of
anthropogenic disturbance and fire ($A_t, F_t$). Recruitment rate $R_t$
is estimated from the observed number of calves $\hat{J}_t$ and adult
female caribou $\hat{W}_t$ assuming a Binomial distribution:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

@@

Following others [@eacker_web-based_2019; @lamb_effectiveness_2024], Kaplan-Meier
estimates [@kaplan_nonparametric_1958] of observed adult female survival probability $\hat{S}_t$ and associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the *survival* R package [@therneau_modeling_2000],
with precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a
Gaussian distributed function of the true survival probability,
truncated between 0 and 1:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

To handle cases in which there is little or no survival data, we use an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed: 
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatIEst)
\end{equation}

@@

For reference, all variables and parameters are defined in supplemental Table A1. Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chains,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

```{r examplePredictionsLowEffort, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamplesLowEffort.png"))
file.rename(here::here("analysis/paper/figs_submit/bayesianModelExamplesLowEffort.pdf"), 
            here::here("analysis/paper/figs_submit/Figure3_bayesianModelExamplesLowEffort.pdf"))
```

```{r figDiffs, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffslowcmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/diffslowcmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure4_diffslowcmult6ay1aSf1.pdf"))
```

```{r figStatusErrors, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/powercmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure5_powercmult6ay1aSf1.pdf"))
```

```{r figEVSI, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/EVSIcmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure6_EVSIcmult6ay1aSf1.pdf"))
```

@@

Figure 4: Posterior predictions from the Bayesian model for population trajectories with informative (a) and weakly informative (b) priors given higher interannual variation ($v_R=0.46,v_S=0.087$) and limited monitoring. The population is monitored for 4 years with 15 collars per year, and 3 cows per collared cow in the composition surveys. See Figure 3 for disturbance trends and other details. The limited local data (open triangles) reduces uncertainty (orange shading) about the observed current state of the population, but is not sufficient to reduce future uncertainty. Given limited data the model with informative priors (a) yields predictions (orange bands) that are consistent with knowledge of observed outcomes across the country (blue shading), and the model with weakly informative priors (b) does not.

@@

Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring and the number of collars reduces uncertainty, but there are diminishing returns on additional monitoring as the number of years increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).

Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$), and monitoring effort. When disturbance is high EVSI is low, indicating that local monitoring does not change predicted status. Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases. Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction. See Figure \@ref(fig:figStatusErrors) for additional scenario details.

>>>>

resultScenario = "S14"
resultSet ="cmult6ay1aSf1"
figSet = c("diffslow","distScns","EVSI","power","EVPIall","powerLam","probViability")

@@

To investigate the effectiveness of various monitoring strategies, and to show how this
varies with landscape condition, we begin by simulating example demographic trajectories and observations of those trajectories. To simulated example trajectories we use a modified version of Johnson et al's [-@johnson_science_2020] demographic model described by Dyson et al. [-@dyson_existing_2022]. For brevity, we refer to this as the "national" model, and for clarity we include a full description in the following paragraphs. Together, a set of example trajectories from the national model represent alternative hypotheses about the true state of a particular caribou population. To assess the accuracy of population status assessments and the expected value of alternative monitoring strategies [@runge_which_2011; @dunham_strategic_2023] we analyze each trajectory of simulated observations using a Bayesian model (section 2.4). We also use the sets of simulated trajectories to confirm that prior predictions from our Bayesian model are consistent with knowledge of the observed distribution of outcomes for caribou across the country [@johnson_science_2020].

@@

In our simulated example populations, mortality risk does not vary over
among seasons or individuals, so the number of collared cows at the time of the composition survey $\hat{T}_t$ depends on the true survival rate $\dot{S}_t$ and the number of collared cows at the beginning of the year $\hat{I}_t$ as follows:
\begin{equation}
\hat{T}_t \sim \text{Binomial}(\hat{I}_t,\dot{S}_t).
  (\#eq:hatI)
\end{equation}

@@

Local demographic and covariate data provided as inputs to the Bayesian
population model include the number of collared cows at the start of the year ($\hat{I}_{t}$), the number of collared animals that survive the year ($\hat{T}_{t}$), the observed number
of adult females and calves in the recruitment surveys
($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per
collared female in the composition survey $w$, and the measures of
anthropogenic disturbance and fire ($A_t, F_t$). Recruitment rate $R_t$
is estimated from the observed number of calves $\hat{J}_t$ and adult
female caribou $\hat{W}_t$ assuming a Binomial distribution:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

@@

Survival $S_t$ is estimated from the observed number of collared cows at the beginning of each year $\hat{I}_t$ and the number of those that survive the year $\hat{T}_t$ assuming a Binomial distribution:
\begin{equation}
\hat{T}_t \sim \text{Binomial}(\hat{I}_t,S_t).
  (\#eq:hatSEst)
\end{equation}

Note that for analysis of real known-fate radio collar data data one should use a survival analysis method that accounts for variation throughout the year in both survival rate and the number of collared animals [e.g. @dalgarno_bboutools_2024; @eacker_web-based_2019; @lamb_effectiveness_2024]. However, our implementation of a two-step non-parametric survival analysis method [following @eacker_web-based_2019] yielded slightly biased survival estimates.

@@

@@

For reference, all variables and parameters are defined in supplemental Table A1. Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chains,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics/tree/EI-paper-submission>.

@@

```{r examplePredictionsLowEffort, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("analysis/paper/figs/bayesianModelExamplesLowEffort.png"))
file.rename(here::here("analysis/paper/figs_submit/bayesianModelExamplesLowEffort.pdf"), 
            here::here("analysis/paper/figs_submit/Figure4_bayesianModelExamplesLowEffort.pdf"))
```

```{r figDiffs, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/diffslowcmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/diffslowcmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure5_diffslowcmult6ay1aSf1.pdf"))
```

```{r figStatusErrors, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powercmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/powercmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure6_powercmult6ay1aSf1.pdf"))
```

```{r figEVSI, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/EVSIcmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/EVSIcmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure7_EVSIcmult6ay1aSf1.pdf"))
```

```{r figStatusErrorsLow, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/powerLamcmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/powerLamcmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure8_powerLamcmult6ay1aSf1.pdf"))
```

```{r figProbViability, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/probViabilitycmult6ay1aSf1.png"))
file.rename(here::here("analysis/paper/figs_submit/powerLamcmult6ay1aSf1.pdf"), 
            here::here("analysis/paper/figs_submit/Figure9_probViabilitycmult6ay1aSf1.pdf"))
```

@@

Figure 4: Posterior predictions from the Bayesian model for population trajectories with informative (a) and weakly informative (b) priors given higher interannual variation ($v_R=0.46,v_S=0.087$) and limited monitoring (4 years, 15 collars per year, 3 cows per collared cow in composition surveys). See Figure 3 for disturbance trends and other details. The limited local data (open triangles) reduces uncertainty (orange shading) about the observed current state of the population, but is not sufficient to reduce future uncertainty. Given limited data the model with informative priors (a) yields predictions (orange bands) that are consistent with knowledge of observed outcomes across the country (blue shading), and the model with weakly informative priors (b) does not.

@@

Figure 6: The probability of a correct status assessment varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios). When disturbance is high the probability of correct status assessment is high even without local monitoring. Where disturbance is low or moderate, increasing the number of years of monitoring increases accuracy, but there are diminishing returns on additional monitoring as the number of years increases. In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year; varying these parameters does not alter qualitative conclusions (Supplements D and E).

Figure 7: The expected value of sample information (EVSI) varies with the amount of anthropogenic disturbance (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort. When disturbance is high EVSI is low, indicating that local monitoring does not change predicted status. Where disturbance is lower, EVSI increases with the number of years of monitoring and the number of collars, but there are diminishing returns on additional monitoring as the number of years increases. Negative EVSI values indicate that the posterior population viability prediction is less certain than the prior prediction. See Figure \@ref(fig:figStatusErrors) for additional scenario details.

Figure 8: In the low anthropogenic disturbance scenario (Figure \@ref(fig:figDisturbance)) probability of a correct status assessment varies with the true growth rate of the population (columns), the number of years projected (rows), the amount of interannual variation (high: $v_R=0.46,v_S=0.087$, low: $v_R=0.23,v_S=0.0435$, none), and monitoring effort (see Figure \@ref(fig:figDisturbance) for scenarios). When anthropogenic disturbance is low and expected to increase (Figure \@ref(fig:figDisturbance)) increasing the number of years of monitoring helps to distinguish populations with current demographic rates substantially lower than 1 (top left), and to identify populations with higher than expected growth rates that fare better as disturbance increases (lower right). Increasing the number of collars more effectively increases near-term accuracy when interannual variation is low (top left). There is little advantage to increasing the number of collars if the goal is long-term projection (bottom row). Increasing the accuracy of status assessment for populations with growth rates close to 1 is more difficult (middle columns). In these examples the composition cow multiplier $w$ is 6 and collars are renewed each year.

Figure 9: The posterior probability that the population is viable ($P(\psi|x,\tau)$ in eq \@ref(eq:EVSI)) predicts the probability of a correct status assessment (estimated from the 114,000 example cases in Figures 5 to 7 with a generalized additive model). Status assessment accuracy increases as the probability of viability nears 0 or 1. Cases in which there is no need for additional monitoring if one is willing to accept a 2% chance of error fall above the dashed line.

### 2025-01-27 analysis/paper/supplement.Rmd
MS revisions in progress

|$\hat{T}_t$|animals|Number of collared cows.|
|$\hat{S}_t$|probability|Kaplan-Meier estimate of observed adult female survival.|
|$\tau_t$||Precision of Kaplan-Meier survival estimate.|
|$\hat{I}_{i,t,m}$|0 or 1|Indicator of whether animal $i$ is dead or alive.|

@@

Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs 4 and \@ref(fig:powercmult6ay1aSf1NoBias)). A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves (supplement B). Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment. Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.

```{r powercmult6ay1aSf1NoBias, fig.cap="The probability of a correct status assessment with unknown bias in recruitment estimates from calf:cow surveys. In these examples there are 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1). See Figure 4 for additional details. ", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult6ay1aSf1.png")))
```

>>>>

|$\hat{T}_t$|animals|Number of collared cows at end of year.|
|$\hat{I}_{t}$|animals|Number of collared cows at beginning of year.|

@@

Adding unknown unknown bias does increase uncertainty, but these increases are not large (compare Figs 6 and \@ref(fig:figStatusErrorsNoBias)). A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults, and the probability of missing calves (supplement B). Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment. Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out.

```{r figDiffsNoBias, fig.cap="The distribution of differences between true expected population growth rates and Bayesian posterior expected growth rates without bias in recruitment estimates from calf:cow surveys. In these examples there is no interannual variation, 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1). See Figure 6 for additional details.", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/diffsnonecmult6ay1aSf1.png")))
```

```{r figStatusErrorsNoBias, fig.cap="The probability of a correct status assessment without bias in recruitment estimates from calf:cow surveys. In these examples there are 6 cows per collared cow in recruitment surveys, and disturbance increases over time (Figure 1). See Figure 6 for additional details. ", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powercmult6ay1aSf1.png")))
```

```{r figStatusErrorsLowNoBias, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/powerLamcmult6ay1aSf1.png")))
```

```{r figProbViabilityNoBias, fig.cap=".", cache=TRUE}
knitr::include_graphics(here::here(paste0("figs/",noBiasResultScenario,"/probViabilitycmult6ay1aSf1.png")))
```
